Black-hole ringdown as a probe of higher-curvature gravity theories

### 2022-03-30 papers/paper_alt_theor_bounds/paper_alt_theor_bounds.tex
hs: moving to the eft sections

In principle, a modification to GR would also affect $M$ and $\chi$ and the expansion should be written
in terms of non-GR mass and spin, say $\bar{M}$ and $\bar{\chi}$. If we assume that the beyond-GR corrections
are included perturbatively, the modifications to the BH mass and spin can be absorved into the deviations
parameters $\delta\omega$ and $\delta\tau$.
%
This means we can identify the $M$ and $\chi$ with their \emph{corresponding GR values}.

We remark that in the GR limit, where $\delta\omega_{k}^{(n)} = \delta\tau_{k}^{(n)} = 0$ $\forall\, n$,
the series~\eqref{eq:kerr_expansion} truncated at $N_{\rm max} = 4$, reproduces with 1\% accuracy
the Kerr QNMs for spins $\chi \lesssim 0.7$.
%
The values of the fitting coefficients $\omega_{k}^{(n)}$ and $\tau_{k}^{(n)}$
can be found in Ref.~\cite{Maselli:2019mjd}, Table I.

% \hs{TODO: in figures and tables use $N_{\rm max} = 0$ ($N_{\rm max} = 1$) instead of $j=0$ ($j=0,1$).}
\hs{TODO: say that we check a posteriori if our PE runs agree with this.}

@@

\begin{figure}[t]
\includegraphics[width=\columnwidth]{figs/edgb_posteriors_combined.pdf}
\caption{(Color online).~Posterior distribution function in EdGB gravity
for the coupling constant $\ell_{\rm GB}$ for GW150914 (top panel), GW200129 (middle panel)
and combined events (bottom panel).
%
In all panels, different line colors correspond to the inclusion ($N_{\rm max} = 1$) or not ($N_{\rm max} = 0$)
of the linear-in-spin QNM correction.}
\label{fig:edgb_exec_sum}
\end{figure}

@@

As we emphasized in Sec.~\ref{sec:remarks}, before drawing any conclusions on
the allowed values for $\ell_{\rm GB}$ from our parameter estimations, we
must first check whether the ``EFT''~\eqref{eq:eft_bound} and
``ParSpec''~\eqref{eq:parspec_bound} bounds are satisfied.
%
We check the validity of the EFT bound in Fig.~\ref{fig:edgb_cdf}. In the top (bottom) panel we show
the cumulative density function of the $\ell_{\rm GB}$ posteriors for GW150914 (GW200129), obtained
evaluating the integral~\eqref{eq:def_cdf} with $\Lambda_{\rm EFT}(\varepsilon, \mathfrak{m})$,
with the mass scale set by the secondary's mass (i.e., $\mathfrak{m} = m_2$, dashed lines) or
the remnant's mass ($\mathfrak{m} = M_f$, solid lines), while varying $\varepsilon$ between 0 and 1.
%
For GW150914, we see that for the $N_{\rm max} = 0$ curves, that the CDF never goes
past $0.2$, regardless of the mass scale $\mathfrak{m}$ used and even at $\varepsilon =
1$. This shows that that the ``EFT bound'' given by Eq.~\eqref{eq:eft_bound} is
never met at a significant credibility and thus that we cannot place a bound on
$\ell_{\rm GB}$.
%
The situation is similar for GW200129 with $N_{\rm max} = 0$ and does not
change for either event when we add spin corrections to the EdGB QNM.
%
Altogether, \emph{we conclude that we cannot constrain EdGB gravity with our
present model}.
%
These conclusions are summarized in Table~\ref{tab:summary_edgb}.

Let us contrast our results with those of Ref.~\cite{Lyu:2022gdr} which placed
the bound
%
$\ell_{\rm GB} \leqslant 1.33$~km
%
using the \emph{inspiral} portion of the neutron star-black hole binaries GW200105 and GW200115~\cite{LIGOScientific:2021qlt} or
of Refs.~\cite{Nair:2019iur,Perkins:2021mhb}, which placed the bound $\ell_{\rm GB} \leqslant 1.7$~km, by considering
a selection of black hole binaries from the GWTC-1 and GWTC-1 catalogs~\cite{}\footnote{These bounds
are, strictly speaking, valid only when the scalar field $\varphi$ is small, i.e., $\varphi \ll 1$, and
take into consideration only the leading-order scalar field contribution arising from the dilatonic coupling.}
%
In this cases

% which relies on post-Newtonian calculations of the GW phase shift in this theory~\cite{Yagi:2011xp,Shiralilou:2020gah,Shiralilou:2021mfl}
% combined with parametrized post-Einsteinian (ppE) framework~\cite{Yunes:2009ke,Chatziioannou:2012rf,Tahura:2019dgr}.

Based on the later, but now here confirmed, within the assumptions of our waveform model,
in the case of comparable mass binaries.

Hence, here we have an example of a theory in which the inspiral portion of the signal
is more constraining then the ringdown portion of the signal.
%
As we will next, the situation is the opposite for dCS gravity.

@@

\begin{figure}[t]
\includegraphics[width=\columnwidth]{figs/dcs_posteriors_combined.pdf}
\caption{(Color online).~Posterior distribution function for the coupling constant $\ell_{\rm CS}$ for
GW150914 (top panel), GW200129 (middle panel) and combined events (bottom panel).
%
In all panels, we use solid and dashed lines to correspond to the inclusion or not of the linear
in spin QNM correction.
%
We stress that posteriors obtained including $j=1$ corrections, violated conditions~\eqref{eq:parspec_bound}
and therefore should not be used to draw meaningful conclusions. We show them for illustrative purposes
and also to emphasize the importance of taking conditions~\eqref{eq:eft_bound} and~\eqref{eq:parspec_bound}
simultaneously into consideration when analyzing the results of the parameter estimation.
}
\label{fig:dCS_exec_sum}
\end{figure}

@@

We now turn our attention to dCS gravity.
%
In Fig.~\ref{fig:dCS_exec_sum} we show the marginalized posterior distribution
function of $\ell_{\rm CS}$ for both events GW150914 (top panel) and GW200129
(middle panel), with (dashed line) and without (solid line) spin corrections to
dominant $l = m = 2$,~$n=0$~QNM.
%
Perhaps the most salient feature of these posteriors is that they are not
peaked at $\ell_{\rm CS}$, which may mislead us to believe that we are seeing a
deviation from GR. This is not the case.
%
We must remember that in the ParSpec formalism that the GR deviations
are controlled by $\gamma \, \delta \omega^{(j)}$~and~$\gamma \, \delta \tau^{(j)}$.
We verified that these quantities do indeed peak at zero indicating consistency with GR.
%
We also see that in both cases, that the inclusion leading-order-in-spin
correction to the mode, displaces the posteriors towards smaller values of
$\ell_{\rm CS}$. This can be seen more evidently by looking at the location of
posterior peaks.
%
Finally, in the bottom panel, we show the combined result for both events.

In Fig.~\ref{fig:dcs_cdf} we show the cumulative distribution function (CDF) as
evaluates with Eq.~\eqref{eq:def_cdf}, with $\ell_{\rm max} = \Lambda_{\rm EFT}$,
for both GW150914 (top panel) and GW200129 (bottom panel), where $\Lambda_{\rm EFT}$
is calculated with either $\mathfrak{m} = m_2$ (blue curves) or $M_{f}$ (orange curves),
and with $\varepsilon \in [0,\, 1]$.
%
For GW150914, we see that with $\gm = m_2$ that Eq.~\eqref{eq:eft_bound} is not
satisfied unless $\varepsilon \approx 0.9$ (with only $j=0$ corrections) and
$\varepsilon \approx 0.7$ (with both $j=0$ and $1$ corrections).
%
The situation is different if we use $\gm = M_f$. In this case, we find that
with or without spin corrections that Eq.~\eqref{eq:eft_bound} can be satisfied
with $\varepsilon \leqslant 1/2$, i.e., below the criteria used
Refs.~\cite{Nair:2019iur,Perkins:2021mhb,Lyu:2022gdr}.
%
This means that with our model's assumptions and using the remnant's source mass $M_f$ to set the
cut-off scale that we can claim an upper bound
%
\begin{equation}
\ell_{\rm CS} \leqslant 42.6~\textrm{km}
% ~( = \Lambda_{\rm EFT}(\tfrac{1}{2}, M_{f}) )
\quad \textrm{(90\% credibility)},
\end{equation}
%
on dCS gravity and would constitute the first bound on this theory with
gravitational-wave observations alone.
%
We also note that since the CDF grows fast in approximate domain $\varepsilon = [0.2, 0.4]$,
that a stronger, albeit at lower credibility,
%
\begin{equation}
\ell_{\rm CS} \leqslant 37.4~\textrm{km}
% ~( = \Lambda_{\rm EFT}(0.3, M_{f}) )
\quad \textrm{at 68\% credibility},
\end{equation}
%
can also be placed, when $N_{\rm max}=0$.

>>>>

In principle, a modification to GR would also affect $M$ and $\chi$ and the expansion should be written
in terms of non-GR mass and spin, say $\bar{M}$ and $\bar{\chi}$. If we assume that the beyond-GR corrections
are included perturbatively, the modifications to the BH mass and spin can be absorbed into the deviations
parameters $\delta\omega$ and $\delta\tau$.
%
This means we can identify the $M$ and $\chi$ with their \emph{corresponding GR values}.
%
\hs{TODO: say that we check a posteriori if our PE runs agree with this.}

We remark that in the GR limit ($\gamma \to 0$)
% where $\delta\omega_{k}^{(n)} = \delta\tau_{k}^{(n)} = 0$ $\forall\, n$,
the series~\eqref{eq:kerr_expansion} truncated at $N_{\rm max} = 4$, reproduces with 1\% accuracy
the Kerr QNMs for spins $\chi \lesssim 0.7$.
%
The values of the fitting coefficients $\omega_{k}^{(n)}$ and $\tau_{k}^{(n)}$
can be found in Ref.~\cite{Maselli:2019mjd}, Table I.

% \hs{TODO: in figures and tables use $N_{\rm max} = 0$ ($N_{\rm max} = 1$) instead of $j=0$ ($j=0,1$).}

@@

\begin{figure}[t]
\includegraphics[width=\columnwidth]{figs/edgb_posteriors_combined.pdf}
\caption{(Color online).~Posterior distribution function in EdGB gravity
for the coupling constant $\ell_{\rm GB}$ for GW150914 (top panel), GW200129 (middle panel)
and combined events (bottom panel).
%
In all panels, different line colors correspond to the inclusion ($N_{\rm max} = 1$) or not ($N_{\rm max} = 0$)
of the linear-in-spin QNM correction.
%
The joint posteriors are shown for illustrative purposes only. As we explain in Fig.~\ref{fig:edgb_cdf} and
in the main text, our analysis of these events fails to satisfy the EFT bound~\eqref{eq:eft_bound}.
}
\label{fig:edgb_exec_sum}
\end{figure}

@@

As we emphasized in Sec.~\ref{sec:remarks}, before drawing any conclusions on
the allowed values for $\ell_{\rm GB}$ from our parameter estimations, we
must first check whether the ``EFT''~\eqref{eq:eft_bound} and
``ParSpec''~\eqref{eq:parspec_bound} bounds are satisfied.
%
We check the validity of the EFT bound in Fig.~\ref{fig:edgb_cdf}. In the top (bottom) panel we show
the cumulative density function of the $\ell_{\rm GB}$ posteriors for GW150914 (GW200129), obtained
evaluating the integral~\eqref{eq:def_cdf} with $\Lambda_{\rm EFT}(\varepsilon, \mathfrak{m})$,
with the mass scale set by the secondary's mass (i.e., $\mathfrak{m} = m_2$, dashed lines) or
the remnant's mass ($\mathfrak{m} = M_f$, solid lines), while varying $\varepsilon$ between 0 and 1.
%
For GW150914, we see that for the $N_{\rm max} = 0$ curves, that the CDF never goes
past $0.2$, regardless of the mass scale $\mathfrak{m}$ used and even at $\varepsilon =
1$. This shows that that the ``EFT bound'' given by Eq.~\eqref{eq:eft_bound} is
never met at a significant credibility and thus that we cannot place a bound on
$\ell_{\rm GB}$.
%
The situation is similar for GW200129 with $N_{\rm max} = 0$ and does not
change for either event when we add spin corrections to the EdGB QNM.
%
Taken together we are led to \emph{conclude that we cannot constrain EdGB gravity with our
present model}.
%
We summarize our findings in Table~\ref{tab:summary_edgb}.

Let us place our results in a broader context. In particular, let us contrast
them against those of Ref.~\cite{Lyu:2022gdr} which placed the bound
%
$\ell_{\rm GB} \leqslant 1.33$~km
%
using the neutron star-black hole binaries GW200105 and GW200115~\cite{LIGOScientific:2021qlt}, and
of Refs.~\cite{Nair:2019iur,Perkins:2021mhb}, which placed the bound
%
$\ell_{\rm GB} \leqslant 1.7$~km,
%
by stacking the posterior of $\ell_{\rm GB}$ from a  selection of black hole binaries from the GWTC-1 and GWTC-2 catalogs
with at least one relatively low-mass binary component~\cite{}\footnote{These bounds
are, strictly speaking, valid only when the scalar field $\varphi$ is small, i.e., $\varphi \ll 1$, and
take into consideration only the leading-order scalar field contribution arising from the dilatonic coupling.}.
%
These works use the \emph{inspiral} portion of GW event only and a constraint
on EdGB gravity can be obtained because black holes in this theory carry a
monopole scalar charge which in turn can source dipolar scalar radiation,
that affects the GW phase at $-1$PN order, relative to the dominant GR contribution, and
is proportional to the difference between the charges of binary components.
%
Now, the magnitude of the scalar charge scales roughly with the inverse of the black hole's mass squared, that is,
smaller (larger) back holes carry large (smaller) scalar charges. Qualitatively, this is because the scalar field
is sourced by the Gauss-Bonnet curvature invariant, which is large for black holes with small masses.
%
Beyond-GR corrections to the inspiral are, by construction, not present in our
waveform model and constitutes a important ingredient to be added in forthcoming
versions of the model.
%
In addition to the larger mass (which suppresses the black hole's charge), the
QNM of EdGB gravity differ very little from GR as observed in~\cite{Blazquez-Salcedo:2016enn}.
%
Hence, here we have an example of a theory in which the inspiral portion of the signal
is more constraining then the ringdown portion of the signal.
%
As we will next, the situation is the opposite for dCS gravity.

@@

\begin{figure}[t]
\includegraphics[width=\columnwidth]{figs/dcs_posteriors_combined.pdf}
\caption{(Color online).~Posterior distribution function for the coupling constant $\ell_{\rm CS}$ for
GW150914 (top panel), GW200129 (middle panel) and combined events (bottom panel).
%
In all panels, we use solid and dashed lines to correspond to the inclusion or not of the linear
in spin QNM correction.
%
We stress that posteriors obtained including $n=1$ corrections, violated conditions~\eqref{eq:parspec_bound}
and therefore should not be used to draw meaningful conclusions. We show them for illustrative purposes
and also to emphasize the importance of taking conditions~\eqref{eq:eft_bound} and~\eqref{eq:parspec_bound}
simultaneously into consideration when analyzing the results of the parameter estimation.
}
\label{fig:dCS_exec_sum}
\end{figure}

@@

We now turn our attention to dCS gravity.
%
In Fig.~\ref{fig:dCS_exec_sum} we show the marginalized posterior distribution
function of $\ell_{\rm CS}$ for both events GW150914 (top panel) and GW200129
(middle panel), with (dashed line) and without (solid line) spin corrections to
dominant $(2,2,0)$~QNM.
%
Perhaps the most salient feature of these posteriors is that they are not
peaked at $\ell_{\rm CS}$, which may mislead us to believe that we are seeing a
deviation from GR. This is not the case and the situation is that same as the
one we discussed in the previous section for EdGB gravity.
%
We verified that
%
\begin{equation}
    \gamma_{\rm CS} = \left( \frac{c^2 \ell_{\rm CS}}{G M_{\rm s}} \right)^{4}\,,
\end{equation}
%
does indeed peak at zero indicating consistency with GR, similarly to what is shown in Fig.~\ref{} for $\gamma_{\rm GB}$.
%
We also see that in both cases, that the inclusion leading-order-in-spin
correction to the mode, displaces the posteriors towards smaller values of
$\ell_{\rm CS}$. This can be seen more evidently by looking at the location of
posterior peaks.
%
Finally, in the bottom panel, we show the combined result for both events.

In Fig.~\ref{fig:dcs_cdf} we show the cumulative distribution function,
obtained from Eq.~\eqref{eq:def_cdf}, with $\ell_{\rm max} = \Lambda_{\rm EFT} (\varepsilon, \gm)$,
for both GW150914 (top panel) and GW200129 (bottom panel), where $\Lambda_{\rm EFT}$
is calculated with either $\gm = m_2$ (dashed lines) or $M_{f}$ (solid lines),
and in the range $\varepsilon \in [0,1]$.
%
For GW150914, we see that with $\gm = m_2$ that Eq.~\eqref{eq:eft_bound} is not
satisfied unless $\varepsilon \approx 0.9$ (with only $n=0$ corrections) and
$\varepsilon \approx 0.7$ (with both $n=0$ and $1$ corrections).
%
The situation is different if we use $\gm = M_f$. In this case, we find that
with or without spin corrections that Eq.~\eqref{eq:eft_bound} can be satisfied
with $\varepsilon \leqslant 1/2$, i.e., below the criteria used
Refs.~\cite{Nair:2019iur,Perkins:2021mhb,Lyu:2022gdr}.
%
This means that with our model's assumptions and using the remnant's source mass $M_f$ to set the
cut-off scale that we can claim an upper bound
%
\begin{equation}
\ell_{\rm CS} \leqslant 42.6~\textrm{km}
\quad \textrm{(90\% credibility)},
\end{equation}
%
on dCS gravity and would constitute the first bound on this theory with
gravitational-wave observations alone.
%
We also note that since the CDF grows fast in approximate domain $\varepsilon = [0.2, 0.4]$,
that a stronger, albeit at lower credibility,
%
\begin{equation}
\ell_{\rm CS} \leqslant 37.4~\textrm{km}
% ~( = \Lambda_{\rm EFT}(0.3, M_{f}) )
\quad \textrm{at 68\% credibility},
\end{equation}
%
can also be placed, when $N_{\rm max}=0$.
